A Tale of Two Transects

### 2019-10-10 writing/transect_paper.Rmd
work on methods section

  - Brendan's pipeline:
      - Ion Torrent
          - used demultiplexing from SciLife (standard IonXpress)
          - primers removed with `cutadapt` [@martin2011]
      - PacBio
          - CCS basecalls re-done with `ccs` version ___ [@pacificbiosciences2019]
          - demultiplexed and primers removed with `cutadapt` [@martin2011]
      - ITS1, ITS2, ITS, LSU extracted using ITSx [@Nilsson2010; @bengtsson-palme2013]
      - extracted regions independently quality filtered by max expected error using `dada2::filterAndTrim()`
      - dereplicated using `dada2::derepFastq()`
      - denoised with `dada2::dada()`
      - de novo chimera removal using `dada2::removeBimeraDenovo()`
      - taxonomy assigned using IDTAXA [@murali2018]
          - UNITE version 8 [@nilsson2019a]
              - extracted ITS1 and ITS2 using ITSx [@Nilsson2010; @bengtsson-palme2013]
              - trained three classifiers: full ITS, ITS1, ITS2.
          - RDP fungal LSU release 11.5 [@cole2014]
              - chose only sequences containing conserved primer sites for ITS4 (3' side of H11; CAUAUnAnUnAGnSSAGGA) and LR5 (on H37; CGAAnUUUCnCUCAGGAUAGCnG).
              - ambiguities based on [@cannone2002; @tedersoo2015]
              - Trim all bases after LR5 because they are not present in our amplicons.
          - SILVA LSU release 132 [@quast2013; @glockner2017; @yilmaz2014]
              - filter out sequences annotates as prokaryotes, then treat as RDP.
          - all databases
              - select medioid sequence for each annotated species, based on minimum 5-mer distance to other sequences
              - for sequences not annotated as species (e.g. "unidentified Basidiomycota"), randomly choose up to 20 per taxon
              - Use up to 5 iterations of IDTAXA training, removing all "problem" sequences on each iteration, unless last member of annotated taxon.
              - "underannotated" sequences (e.g., "unidentified Basidiomycota" confidently placed as Agaricomycetes) not protected from removal if last sequence annotated as such.
      - ecological guild assigned using FUNGuild [@nguyen2016funguild]
      - consensus long read and LSU region constructed for each ITS2 ASV:
          - align reads using `DECIPHER` with RNA secondary structure calculation [@wright2015;@wright2016]
          - remove outliers (i.e. chimeras) with `odseq` [@jehl2015]
          - remove gap columns (>50%)
          - take consensus (>50%) at non-gaps
          - remove sequences with more than 3 ambiguous bases (bad consensus)
      - initial alignment of long amplicons with DECIPHER [@wright2015;@wright2016]
      - alignment/tree for long amplicons built iteratively with PASTA[@mirarab2014; @mirarab2014a]. Fastree tree, mafft align, OPAL merge, RaxML final tree

>>>>

  - Brendan's pipeline:
      - Ion Torrent
          - used demultiplexing from SciLife (standard IonXpress)
          - primers removed with `cutadapt` [@martin2011]
      - PacBio
          - CCS basecalls re-done with `ccs` version ___ [@pacificbiosciences2019]
          - demultiplexed and primers removed with `cutadapt` [@martin2011]
      - Illumina MiSeq
          - pending
      - Region extraction with LSUx (available as R package on github and conda)
          - 5.8S located with Rfam [@kalvari2018] covariance model (CM) using `cmsearch` from Infernal 1.1.2 [@nawrocki2013]
          - read trimmed at 5' end of 5.8S; remaining read aligned to modified 5.8S + LSU (=32S) CM from Ribosomal Data Project (RDP) [@glockner2017, @cole2014] with annotations for different LSU domains (definitions from @michot1984 and @raue1988), using `cmalign` from Infernal 1.1.2
          - 5.8S, ITS2, D1, D2, D3, and conserved LSU domains (called LSU1, LSU2, LSU3, LSU4) extracted from alignment using annotations; ITS1 is everything before 5.8S.
      - extracted regions independently quality filtered by max expected error using `dada2::filterAndTrim()`
      - dereplicated using `dada2::derepFastq()`
      - denoised with `dada2::dada()`
      - de novo chimera removal using `dada2::removeBimeraDenovo()`
      - full length amplicons reconstructed from ASVs of extracted domains (r package tzara -- available on github and conda)
          - when a read is mapped to ASVs for all regions, they are concatenated.
          - bimera detection on each trio of variable-conserved-variable domains (e.g., ITS1-5.8S-ITS2, ITS2-LSU1-D1, etc.); reads which are detected at bimera for any trio are discarded.
          - For each ITS2 ASV, check each other domain:
              - If some reads are mapped to ASVs, but some are not, assign the unassigned to the closest ASV as long as distance < 10. This is less stringent than DADA2.
              - If no reads are mapped to ASVs, calculate consensus sequence:
                  - align reads using `DECIPHER` with RNA secondary structure calculation [@wright2015;@wright2016]
                  - remove outliers (i.e. chimeras) with `odseq` [@jehl2015]
                  - remove gap columns (>50%)
                  - take consensus (>50%) at non-gaps
                  - remove sequences with more than 3 ambiguous bases (bad consensus)
          - repeat process using each other domain as reference, in decreasing order of diversity.

      - taxonomy assigned using IDTAXA [@murali2018]
          - UNITE version 8 [@nilsson2019a]
              - extracted ITS1 and ITS2 using ITSx [@Nilsson2010; @bengtsson-palme2013]
              - trained three classifiers: full ITS, ITS1, ITS2.
          - RDP fungal LSU release 11.5 [@cole2014]
              - chose only sequences containing conserved primer sites for ITS4 (3' side of H11; CAUAUnAnUnAGnSSAGGA) and LR5 (on H37; CGAAnUUUCnCUCAGGAUAGCnG).
              - ambiguities based on [@cannone2002; @tedersoo2015]
              - Trim all bases after LR5 because they are not present in our amplicons.
          - SILVA LSU release 132 [@quast2013; @glockner2017; @yilmaz2014]
              - filter out sequences annotates as prokaryotes, then treat as RDP.
          - all databases
              - select medioid sequence for each annotated species, based on minimum 5-mer distance to other sequences
              - for sequences not annotated as species (e.g. "unidentified Basidiomycota"), randomly choose up to 20 per taxon
              - Use up to 5 iterations of IDTAXA training, removing all "problem" sequences on each iteration, unless last member of annotated taxon.
              - "underannotated" sequences (e.g., "unidentified Basidiomycota" confidently placed as Agaricomycetes) not protected from removal if last sequence annotated as such.
      - ecological guild assigned using FUNGuild [@nguyen2016funguild]
      - consensus long read and LSU region constructed for each ITS2 ASV:

      - initial alignment of long amplicons with DECIPHER [@wright2015;@wright2016]
      - alignment/tree for long amplicons built iteratively with PASTA[@mirarab2014; @mirarab2014a]. Fastree tree, mafft align, OPAL merge, RaxML final tree
